---
title: "Start/stop rules in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

#### Why do we want highest `streak_1 >= x`, and not highest `streak_1 = x`?

To rephrase, we are looking for all `streak_1` of length _x_ or greater, not only `streak_1` whose length is exactly _x_. To see why, consider cases `A` and `B` below.

```{r tall_cases, eval=FALSE}
 item      A     B
 i01       1     1
 i02       1     1
 i03       0     0
 i04       1     1
 i05       1     1
 i06       1     1
 i07       1     1
 i08       0     0
 i09       0     0
 i10       1     1
 i11       0     0
 i12       1     1
 i13       0     1
 i14       0     1
 i15       1     1
 i16       1     1
 i17       1     1
 i18       1     1
 i19       0     0
 i20       1     1
```
Both cases have first `streak_1` of length 4. The difference between the two is that `A` has a high `streak_1` of 4, beginning at `i15`, whereas `B` has a high `streak_1` of 7, beginning at `i12`. The importance of these high `streak_1` is that they determine where we set the start item for each case.

Intuitively, we can see that `B` represents higher ability, simply because it yields a higher raw score on the test. And, start items are set higher in the test as a function of increasing ability.

Now, imagine that we set start item by looking for the highest `streak_1` whose length is exactly _x_. For `A`, that `streak_1` begins at `i15` (as noted above). But for `B`, that `streak_1` begins at `i04`.

This is because our search rule (streak_length of exactly _x_) specifies that we must ignore the streak of 1s that begins at `i12`, becaues its length is not exactly _x_. This is counterintuitive, because even though its overall length is longer than _x_, it contains several streaks whose length is exactly _x_.

A further counterintuitive result of the exactly _x_ search rule is that it causes the start item of `B` to be set lower than that of `A`. This is counterintuitive because we already agreed that `B` has higher ability than `A`, and therefore should have a higher start item than `A`.

These problems are solved by finding the highest `streak_1 >=x`. This search rule catches `B`'s `streak_1` of length 7 that begins at `i12`, because that streak contains a `streak_1` of length 4. Thus, the start item will be set within that high streak, as is appropriate for `B`'s ability.
